He-Man: A domain-specific language for lightweight concurrency

### 2011-12-11 report/report.tex
Hacked out report introduction.

\documentclass{article}
\usepackage{cite}
\usepackage[colorlinks=true,linkcolor=black,urlcolor=black,citecolor=black]%
{hyperref}
\renewcommand{\t}{\texttt}

@@

Modern network applications need to support many simultaneous clients---this is
known as the C10K (``10,000 client'') problem\cite{Kegel}. Modern servers have
enough memory and network resources to support this many

To maintain these simultaneous connections, an application must maintain each
client's state independently;

while 
block each one

The simplest way to do this is to spawn a thread for each client, but this
approach does not scale as the number of clients grows.

A lightweight approach is to use kernel mechanisms for multiplexed, non-blocking
I/O, such as \t{select}, \t{poll}, \t{epoll}, or \t{kquery}; unfortunately, to
maintain separate state for each client, application programmers must manually
structure their code as state machines, which is tedious and error-prone.
Unsurprisingly, there exist many thin C/C++ wrappers for these facilities, but
none avoid major code refactoring.

To maintain the illusion of threads (in particular, client-specific state),
languages like Erlang and Haskell provide lightweight threading via user-space
schedulers and non-blocking IO. While this affords better performance, the full
generality of this approach still imposes unnecessary overhead on networking
applications which don't require preemptable, independently scheduled threads
\cite{Vinoski}.

Because

Hence, we intend to produce an embedded domain-specific language in Haskell to
describe concurrent network applications. This language will provide composable,
high-level network operations, in which programmers may express their
application logic naturally, as if using threads.

However, this high-level description will be automatically synthesized into a
state machine representation usable in a C loop invoking \t{epoll}, using no
threads whatsoever. We will additionally output a graphical representation of
this underlying state machine.

Generate C source code which itself implements the network application, linkable
with native C code.

simple runtime

@@

The Nginx\cite{Nginx} and Lighttpd\cite{Lighttpd} HTTP servers use an
event-based architectures, unlike Apache's threads\cite{Apache}, with the
specific goal of addressing the C10k problem,

@@

Haskell has epoll support and lightweight threads

Erlang

@@

@@

>>>>

\documentclass[twocolumn]{article}
\usepackage{cite}
\usepackage[colorlinks=true,linkcolor=black,urlcolor=black,citecolor=black]%
{hyperref}
\renewcommand{\t}{\texttt}
\renewcommand{\b}{\textbf}
\renewcommand{\i}{\textit}

\newcommand{\prettybox}[1]{\fbox{\parbox{\linewidth}{#1}}}

% A bunch of crazy BNF TeX I wrote once
\def\bnf#1{\prettybox{\halign{##\hfil& \hfil ##\hfil& \ ##\hfil\cr #1}}}
\def\p#1{& = & #1 \cr}
\def\pp#1{& & #1 \cr}
\def\a#1{& $\mid$ & #1 \cr}

@@

Network applications need to support many simultaneous clients---this is known
as the C10K (``10,000 client'') problem\cite{Kegel}. Modern servers have enough
memory and bandwidth for massive concurrency, but it is difficult to design
applications to take full advantage of these resources.

A server must maintain separate state for each connection, and often waits for
network or disk I/O to complete before continuing to serve a particular client.
The simplest solution is to maintain one thread per client, using blocking I/O
within each thread. This incurs frequent and expensive context switches at the
operating system level; because the threads have identical code and little
private data, this approach is unnecessary and does not scale.

An improvement is to write multithreaded code, in which each thread serves
multiple clients. This is the most popular approach, and is utilized in the
Apache HTTP server \cite{Apache}.

Languages like Erlang and Haskell provide lightweight threads via user-space
schedulers. This affords even better performance, but remains unnecessarily
powerful for this problem space, imposing runtime overhead on applications which
don't require preemptable, independently scheduled threads \cite{Vinoski}.

By far the leanest approach is to write an event loop around kernel facilities
for multiplexed, non-blocking I/O, such as \t{select}, \t{poll}, \t{epoll}, or
\t{kquery}. In this model, all I/O occurs at a single call site in the code.
The programmer is responsible for explicitly maintaining per-client state and
invoking the appropriate client on each I/O event.

This requires the entire application to be restructured as a state machine with
explicit per-client control blocks, a model which is very difficult for
programmers to reason about. There exist numerous C/C++ libraries for event
loops, but none of these avoid the state machine paradigm.

Our observation is that the event loop model, in programming languages parlance,
is essentially trampolining continuation-passing-style, which can be obtained by
mechanical translation from the straight-line threaded code. This suggests it is
possible to obtain the best of both worlds---a threaded programming model with
the performance of event loops.

To that end, we present He-Man: Haskell Event Manager, Apropos Networking, an
embedded domain-specific language (EDSL) hosted in Haskell, which provides a
threaded abstraction for network application programming and compiles to a C
event loop based around Linux's \t{epoll}.

He-Man is a straightforward monadic language with built-in network primitives,
offering a simple thread-per-client programming model. It is easily extensible
in Haskell, and also allows calls to foreign C functions; it outputs C code
which links with an event loop runtime and any code a programmer might prefer to
write directly in C. Our vision is that C programmers can use He-Man for their
core application logic, and can write the rest of their application in C.

@@

The Lighttpd\cite{Lighttpd} and Nginx\cite{Nginx} HTTP servers use event-based
architectures, with the stated goal of addressing the C10k problem. Indeed, they
are fast (XXX CITATION)

@@

As mentioned in the intro,
Haskell has epoll support and lightweight threads
Erlang

@@

The back-end converts our linear front-end code into a collection of states

is to compile C,

It's important that the only blocking operations happen in the main loop

We separate

Threads are separated and replaced by labels

pass to make it

\begin{figure}[ht]
\centering
\bnf{
\i{Thread}
\p{(\i{thread\_label}, [(\i{Var}, \i{Type})])}
\i{Block}
\p{(\i{label}, \i{thread\_label}, [\i{Stmt}], \i{Tail})}
\i{Tail}
\p{\b{Goto} \i{label}}
\a{\b{GotoWait} \i{label}}
\a{\b{If} \i{Expr} \i{Tail} \i{Tail}}
\a{\b{Exit}}
\i{Stmt}
\p{\b{Decl} (\i{Var}, \i{Type}) \i{Expr}}
\a{\b{Assign} \i{Expr} \i{Expr}}
\a{\b{Exp} \i{Expr}}
\a{\b{Spawn} \i{label} [((\i{Var}, \i{Type}), \i{Expr})]}
}
\caption{Back-end grammar}
\end{figure}

@@

write a better allocator

We will additionally output a graphical representation of this underlying state
machine.
